---
title: STAT 4194 Project
author:
- Andre Contreras
- Jiayi Gao
- Mitch Krenz
- Paarth Mahajan
- Tony Zheng
format:
    typst:
        mainfont: calibri
        df-print: paged
    revealjs: default
execute:
    include: false
    cache: true
    autodep: true
---

## Packages Used

```{r}
#| label: libraries
library(tidyverse)
library(igraph)
library(rgexf)
```

# Data


# Model Fits

## AME Model

For the AME model, we only use the sociality and multiplicative effects, since there are no dyadic or nodal attributes for the characters.
```{r}
#| label: amen library
library(amen)
```

Loading movies dataset
```{r}
#| label: read movies, id_by_title
movies_metadata <- read_tsv("network_metadata.tab")

get_movie_id_by_title <- function(title) {
    movies_metadata |>
        filter(grepl(title, Title)) |>
        pull(GexfID)
}
```

# Movie Specific Examples

```{r}
#| label: model_movie_example
# creates some exmaple plots and GOF for a movie
model_movie_example <- function(title, cache = TRUE) {
    movie_id <- get_movie_id_by_title(title)
    cache_file <- str_glue("cache/{movie_id}_ame_model")
    if (cache && file.exists(cache_file)) {
        return(readRDS(cache_file))
    }

    movie_graph <- read.gexf(str_glue("data/{movie_id}.gexf")) |> gexf.to.igraph()

    Y <- movie_graph |>
        laplacian_matrix(sparse = FALSE)
    Y <- -Y
    diag(Y) <- NA

    model <- ame(
        Y = Y,
        family = "frn",
        R = 2,
        nvar = TRUE,
        odmax = E(movie_graph)$weight |> max(),
        symmetric = TRUE,
        plot = FALSE,
        print = FALSE,
        gof = TRUE,
    )

    out <- list(movie_graph = movie_graph, model = model)
    if (!dir.exists("cache")) dir.create("cache")
    saveRDS(out, cache_file)
    return(out)
}

read_model_movie_example <- function(title) {
    cache_file <- str_glue("cache/{movie_id}_ame_model")
    return(readRDS(cache_file))
}
```

## 10 Things I Hate About You

**Kat, Patrick** - Primary love interests

**Bianca, Cameron** - Secondary love interests

**Joey** - Main antagonist


## Model Plot

```{r}
#| label: 10 things I hate about you
ten_things <- model_movie_example("10 Things")
```

```{r, include=TRUE}
#| label: plot 10 things model
plot(ten_things$model)
par(mfrow = c(1, 1))
```

## Movie Plot

```{r, include=TRUE}
#| label: plot 10 things graph
#| fig-width: 16
#| fig-height: 9
vertex.strength <- strength(ten_things$movie_graph)
plot(
    ten_things$movie_graph,
    edge.width = E(ten_things$movie_graph)$weight,
    vertex.size = vertex.strength / max(vertex.strength) * 20 + 10,
    layout = ten_things$model$U,
    vertex.label.font = 2,
    vertex.label.cex = log(vertex.strength, 5) / max(log(vertex.strength, 3)) * 2,
    vertex.label.family = "sans",
    vertex.label.color = "black",
    mark.expand = 0,
    mark.border = NA,
)
```

## Star Wars

**Main characters:** Threepio, Leia, Han, Luke

**Main Antagonists:** Darth Vader, Boba Fett, Emperor

## Model Plot

```{r}
#| label: star wars
star_wars <- model_movie_example("Star Wars.*The Empire Strikes Back")
```

```{r, include=TRUE}
#| label: plot star wars model
plot(star_wars$model)
par(mfrow = c(1, 1))
```

## Movie Plot

```{r, include=TRUE}
#| label: plot star wars graph
#| fig-width: 16
#| fig-height: 9
vertex.strength <- strength(star_wars$movie_graph)
plot(
    star_wars$movie_graph,
    edge.width = E(star_wars$movie_graph)$weight,
    vertex.size = vertex.strength / max(vertex.strength) * 20 + 10,
    layout = star_wars$model$U,
    vertex.label.font = 2,
    vertex.label.cex = log(vertex.strength, 5) / max(log(vertex.strength, 3)) * 2,
    vertex.label.family = "sans",
    vertex.label.color = "black",
    mark.expand = 0,
    mark.border = NA,
)
```

# General Results

## Random Movie Sample

Took random sample of 20 movies and ran same procedure...
```{r, include=TRUE}
#| label: read random movies sample
movies_sample <- read_tsv("488	The Last Samurai\n682	Reservoir Dogs\n130	Black Rain\n134	Blade Runner\n671	The Reader\n843	TRON\n495	Leviathan\n41	Alien: Resurrection\n816	Thirteen Days\n227	The Crying Game\n511	Lord of Illusions\n791	Sugar & Spice\n487	The Last of the Mohicans\n217	Crank\n116	Being Human\n471	The King's Speech\n189	Chasing Sleep\n244	Death to Smoochy\n320	French Kiss\n214	Coraline", col_names = c("id", "title"))

movies_sample$title
```

```{r}
#| label: loop over sample, write each
for (movie_num in 1:nrow(movies_sample)) {
    title <- movies_sample[[movie_num, 2]]
    # for (movie_num in 1:nrow(movies_sample)) {
    model_movie_example(title)
}
```