---
title: STAT 4194 Project
subtitle: Erdos-Renyi Model
author:
- Andre Contreras
- Jiayi Gao
- Mitch Krenz
- Paarth Mahajan
- Tony Zheng
format:
    typst:
        mainfont: calibri
        df-print: paged
    revealjs: default
execute:
    echo: false
    cache: true
    autodep: true
---

## Packages Used
```{r, message=FALSE}
#| label: libraries
library(tidyverse)
library(igraph)
library(rgexf)

# Data Setup
movie_ids <- c(488, 682, 130, 134, 671, 843, 495, 41, 816, 227,
               511, 791, 487, 217, 116, 471, 189, 244, 320, 214)

movie_names <- c(
  "488" = "The Last Samurai", "682" = "Reservoir Dogs", "130" = "Black Rain",
  "134" = "Blade Runner", "671" = "The Reader", "843" = "TRON",
  "495" = "Leviathan", "41"  = "Alien: Resurrection", "816" = "Thirteen Days",
  "227" = "The Crying Game", "511" = "Lord of Illusions", "791" = "Sugar & Spice",
  "487" = "The Last of the Mohicans", "217" = "Crank", "116" = "Being Human",
  "471" = "The King's Speech", "189" = "Chasing Sleep", "244" = "Death to Smoochy",
  "320" = "French Kiss", "214" = "Coraline"
)
```


## Top Partnerships in Each Movie
Before diving into statistical modeling, I first examined the raw structure of the character networks to identify who drives the narrative. In a movie script, character interactions are rarely equal. They are usually centered around a few key relationships. I will take a subset of 20 random movies to look at who their top character partnership was and how many scenes this duo shot together.\
```{r}
# Initialize an empty list to store row data
table_data <- list()

for (id in as.character(movie_ids)) {
  f <- paste0("data/", id, ".gexf")
  
  # Ensure the file exists before processing
  if (!file.exists(f)) next
  
  # Import and clean
  g <- read.gexf(f) |> gexf.to.igraph()
  # Simplify removes self-loops but keeps the scene weights on edges
  g <- simplify(g, remove.multiple = FALSE, remove.loops = TRUE)
  
  # Extract the edge with the highest weight (Max Scenes)
  edge_df <- as_data_frame(g, what="edges")
  top_row <- edge_df %>% arrange(desc(weight)) %>% head(1)
  
  # Store results
  table_data[[id]] <- data.frame(
    Movie = movie_names[id],
    Nodes = vcount(g),
    Edges = ecount(g),
    Top_Partnership = paste(top_row$from, "&", top_row$to),
    Max_Scenes = top_row$weight
  )
}

movie_summary_table <- bind_rows(table_data)
print(movie_summary_table)
```

By looking at the `Max_Scenes` column, I can see that Bertie & Lionel in The King's Speech act as massive anchors for the entire network. This degree heterogeneity is my first hint that a simple random model, like the Erdos-Renyi model where everyone has roughly the same number of friends, will almost certainly fail to capture the reality of these scripts.

## The Erdos-Renyi (ER) Model
To test if cinematic social structures are just a result of random chance, I applied the Erdos-Renyi (ER) method. This model assumes that an edge between any two characters is formed with a fixed probability $p$, independent of any other edges. I estimated $\hat{p}$ for each movie based on its density to create 100 simulated "random twins" for each film.

### Summary Statistics of ER Model vs Observed
```{r}
# ER Analysis Loop
er_results <- data.frame()
sim_dist_example <- list() 

for (id in as.character(movie_ids)) {
  f <- paste0("data/", id, ".gexf")
  if (!file.exists(f)) next
  
  g_real <- read.gexf(f) |> gexf.to.igraph()
  g_real <- simplify(g_real, remove.multiple = FALSE, remove.loops = TRUE)
  
  # ER Parameters & Simulation 
  n <- vcount(g_real)
  phat <- ecount(g_real) / choose(n, 2)
  
  # Running a small batch of simulations to get a distribution 
  b <- 100
  c_star <- replicate(b, transitivity(sample_gnp(n, phat), type="global"))
  m_star <- replicate(b, modularity(cluster_louvain(sample_gnp(n, phat))))
  
  # Calculated Observed
  obs_c <- transitivity(g_real, type="global")
  obs_m <- modularity(cluster_louvain(g_real))
  
  er_results <- rbind(er_results, data.frame(
    movie_name = movie_names[id],
    n.nodes = n,
    p.hat = round(phat, 4),
    obs.clus = round(obs_c, 4),
    er.clus = round(mean(c_star), 4),
    obs.mod = round(obs_m, 4),
    er.mod = round(mean(m_star), 4)
  ))
  
  # Reservoir Dogs distribution for the visual
  if(id == "682") sim_dist_example <- list(star = c_star, obs = obs_c)
}

print(er_results)
```

### Analysis of ER Model
My results highlight a massive structural difference between real films and the ER simulations. In the table above, the **observed clustering** is consistently two to five times higher than the **ER clustering**. In a movie like Reservoir Dogs, characters form triangles (if Character A knows B and C, B and C almost certainly know each other). The ER model completely misses this, as it doesn't account for the cliqueish nature of social interactions.\

### Small World?
```{r}
obs_L <- mean_distance(g_real)
er_L  <- mean_distance(sample_gnp(n, phat))

small_world_summary <- tibble(
  Metric = c("Clustering Coefficient (C)", "Average Path Length (L)"),
  Observed = c(obs_c, obs_L),
  Simulated_ER = c(mean(c_star), er_L)
)

small_world_summary
# Small world check: Is (obs_c/er_c) large AND (obs_L/er_L) close to 1?
```

**Note: A network is small world if Observed C » C_R & L is approximately the same as L_R!**\
Since the clustering is significantly higher but the path length remains relatively equally short, I can classify this movie network as a Small-World Network.\

## Density Plot: The Clustering Gap Between Observed & ER Simulation
```{r}
#| echo: false
#| fig-height: 4
plot(density(sim_dist_example$star), 
     main = "Clustering: Reservoir Dogs vs ER",
     xlab = "Global Clustering Coefficient", 
     col = "blue", lwd = 2,
     xlim = c(0, max(sim_dist_example$obs, sim_dist_example$star) + 0.1))
abline(v = sim_dist_example$obs, col = "darkred", lwd = 2, lty = 2)
legend("topleft", legend = c("Simulated ER", "Observed Movie"), 
       col = c("blue", "darkred"), lty = c(1, 2), bty = "n")
```
The blue curve represents the "random world." The dashed red line represents the real movie. Because the red line is sitting so far to the right, I can statistically conclude that the ER model is a bad fit. It cannot create the community structure that exists in a real script.

## Visualizing the Network Structure of Reservoir Dogs Characters
Finally, I plotted the real character network of Reservoir Dogs alongside an ER simulation of the same size and density. I used the Louvain algorithm to detect and color code communities.
```{r}
# Visualizing the "Hairball" (ER) vs. Narrative Structure (Real)
# Using Reservoir Dogs for this example
set.seed(4194)
target_id <- "682"
g_real <- read.gexf(paste0("data/", target_id, ".gexf")) |> gexf.to.igraph()
g_real <- simplify(g_real)

# Generate ER equivalent
n <- vcount(g_real)
phat <- ecount(g_real) / choose(n, 2)
g_er <- sample_gnp(n, phat)

# Community Detection for coloring
cl_real <- cluster_louvain(g_real)
cl_er <- cluster_louvain(g_er)

# Plotting Side-by-Side
par(mfrow=c(1,2), mar=c(1,1,3,1))

# Real Network
plot(g_real, 
     layout=layout_with_fr(g_real), 
     vertex.color=membership(cl_real),
     vertex.size=10, vertex.label=NA,
     main="Real Movie: Organized Groups")

# ER Network
plot(g_er, 
     layout=layout_with_fr(g_er), 
     vertex.color=membership(cl_er),
     vertex.size=10, vertex.label=NA,
     main="ER Model: Randomness")
```

In the Real Movie plot, we see clear clumps of color representing characters who are forced to interact by the plot. In the ER Model plot, the network looks like a uniform "hairball". While the algorithm still tries to assign colors, the groups are disorganized because the edges were placed by a random coin flip rather than a storyteller. This confirms that while ER models are useful for understanding simple connectivity, they fail to capture the intentional, clustered design of human narratives.\
\
As a result, we can move on to test if other more advanced models we learned in class can be more accurate in detecting relationships between characters.\
