---
title: STAT 4194 Project
author:
- Andre Contreras
- Jiayi Gao
- Mitch Krenz
- Paarth Mahajan
- Tony Zheng
format:
    typst:
        mainfont: calibri
        df-print: paged
    revealjs: default
execute:
    echo: false
    cache: true
    autodep: true
---

## Packages Used

```{r}
#| label: libraries
library(tidyverse)
library(igraph)
library(rgexf)
```

# Data


# Model Fits

## ER Model

```{r, message=FALSE, echo=FALSE}
# Data Setup
dir.create("data", showWarnings = FALSE)
base_url <- "https://raw.githubusercontent.com/tzheng7679/STAT-4194-Project/main/data/"

# Specific sample of 20 movies
movies_sample <- read_tsv(I("id\ttitle\n488\tThe Last Samurai\n682\tReservoir Dogs\n130\tBlack Rain\n134\tBlade Runner\n671\tThe Reader\n843\tTRON\n495\tLeviathan\n41\tAlien: Resurrection\n816\tThirteen Days\n227\tThe Crying Game\n511\tLord of Illusions\n791\tSugar & Spice\n487\tThe Last of the Mohicans\n217\tCrank\n116\tBeing Human\n471\tThe King's Speech\n189\tChasing Sleep\n244\tDeath to Smoochy\n320\tFrench Kiss\n214\tCoraline"))

er_analysis_results <- list()

# Looping over the sample
for(i in 1:nrow(movies_sample)) {
  movie_id <- movies_sample$id[i]
  movie_title <- movies_sample$title[i]
  dest_file <- str_glue("data/{movie_id}.gexf")
  
  # Download check
  if (!file.exists(dest_file)) {
    url <- paste0(base_url, movie_id, ".gexf")
    try(download.file(url, dest_file, mode = "wb", quiet = TRUE))
  }

  # Importing and converting to igraph
  g_real <- tryCatch({
    read.gexf(dest_file) |> gexf.to.igraph()
  }, error = function(e) return(NULL))
  
  if (is.null(g_real)) next
  
  # Clean: remove self-loops but keep multiple edges for partnership weights
  g_real <- simplify(g_real, remove.multiple = FALSE, remove.loops = TRUE)

  # Analysis: Finding Top Partnerships (On-screen scenes)
  edge_list <- as_data_frame(g_real, what="edges")
  top_partnership <- edge_list %>% arrange(desc(weight)) %>% head(1)
  
  # Analysis: ER Model Parameters
  n <- vcount(g_real)
  m <- ecount(g_real)
  p_hat <- m / choose(n, 2)  # Density
  
  # Analysis: Simulating ER Null Model
  g_er <- sample_gnp(n, p_hat)
  
  # Comparing Metrics
  obs_trans <- transitivity(g_real, type="global")
  er_trans <- transitivity(g_er, type="global")
  
  obs_mod <- modularity(cluster_louvain(g_real))
  er_mod <- modularity(cluster_louvain(g_er))
  
  # Storing results
  er_analysis_results[[movie_title]] <- data.frame(
    Title = movie_title,
    Nodes = n,
    Edges = m,
    P_Estimate = round(p_hat, 4),
    Obs_Clustering = round(obs_trans, 4),
    ER_Clustering = round(er_trans, 4),
    Obs_Modularity = round(obs_mod, 4),
    ER_Modularity = round(er_mod, 4),
    Top_Partnership = paste(top_partnership$from, "&", top_partnership$to),
    Max_Scenes = top_partnership$weight
  )
}

final_er_table <- bind_rows(er_analysis_results)
print(final_er_table)
```
To analyze the movie character networks, I imported the co-appearance data for 20 films, treating each character as a node and their shared scenes as weighted edges. I then calculated the density ($\hat{p}$) of each real network to establish a baseline for the Erdos-Rényi (ER) model, which assumes that every relationship forms with equal, independent probability. By simulating these random "null" graphs, we compared the **observed** structural metrics, specifically transitivity and modularity, against the values expected by pure chance to determine if cinematic social structures follow a random or intentional design.

### Analysis of ER Model
The data reveals a massive clustering gap between the real observed movies and my simulation. In the actual films, the `Obs_Clustering` is consistently two to three times higher than the `ER_Clustering`. This shows that movie characters don't just meet at random. They form tight "cliques" where a friend of a friend is almost always your friend too.\
\
My simulation would also theoretically struggle to replicate the     `Top_Partnership` and `Max_scenes` columns. The one's in the table are data from the observed data. In a simulation, all characters would have the same number of co-appearances with the same weight.\

## AME Model

For the AME model, we only use the sociality and multiplicative effects (R = 2), since there are no dyadic or nodal attributes for the characters.

## AME Model Type
We also used a fixed rank nomination model, to fit for the count adjacency data (# of scenes shared). For example
```{r}
#| label: adjacency matrix example
movie_graph <- read.gexf(str_glue("data/1.gexf")) |> gexf.to.igraph()
Y <- movie_graph |>
    laplacian_matrix(sparse = FALSE)
Y <- -Y
diag(Y) <- NA

Y
```

<!-- function defs -->
```{r}
#| label: read movies, id_by_title
library(amen)

movies_metadata <- read_tsv("network_metadata.tab")

get_movie_id_by_title <- function(title) {
    movies_metadata |>
        filter(grepl(title, Title)) |>
        pull(GexfID)
}
```

```{r}
#| label: model_movie_example
# creates some exmaple plots and GOF for a movie
model_movie_example <- function(title, cache = TRUE) {
    movie_id <- get_movie_id_by_title(title)
    cache_file <- str_glue("cache/{movie_id}_ame_model.rds")
    if (cache && file.exists(cache_file)) {
        return(readRDS(cache_file))
    }

    movie_graph <- read.gexf(str_glue("data/{movie_id}.gexf")) |> gexf.to.igraph()

    Y <- movie_graph |>
        laplacian_matrix(sparse = FALSE)
    Y <- -Y
    diag(Y) <- NA

    model <- ame(
        Y = Y,
        family = "frn",
        R = 2,
        nvar = TRUE,
        odmax = E(movie_graph)$weight |> max(),
        symmetric = TRUE,
        plot = FALSE,
        print = FALSE,
        gof = TRUE,
    )

    out <- list(movie_graph = movie_graph, model = model)
    if (!dir.exists("cache")) dir.create("cache")
    saveRDS(out, cache_file)
    return(out)
}

read_model_movie_example <- function(title) {
    cache_file <- str_glue("cache/{movie_id}_ame_model.rds")
    return(readRDS(cache_file))
}
```

# Movie Specific Examples

## 10 Things I Hate About You

**Kat, Patrick** - Primary love interests

**Bianca, Cameron** - Secondary love interests

**Joey** - Main antagonist


## Model Plot

```{r}
#| label: 10 things I hate about you
ten_things <- model_movie_example("10 Things")
plot(ten_things$model)
par(mfrow = c(1, 1))
```

## Latent Positions Plot

```{r}
#| label: plot 10 things graph
#| fig-width: 16
#| fig-height: 9
vertex.strength <- strength(ten_things$movie_graph)
plot(
    ten_things$movie_graph,
    edge.width = E(ten_things$movie_graph)$weight,
    vertex.size = vertex.strength / max(vertex.strength) * 20 + 10,
    layout = ten_things$model$U,
    vertex.label = stringr::str_replace_all(
        vertex_attr(ten_things$movie_graph)$name,
        "(\\w) (\\w)", 
        "\\1\n\\2"
    ),
    vertex.label.font = 2,
    vertex.label.cex = log(vertex.strength, 5) / max(log(vertex.strength, 3)) * 2,
    vertex.label.family = "sans",
    vertex.label.color = "black",
    mark.expand = 0,
    mark.border = NA,
)
```

## Star Wars

**Main characters:** Threepio, Leia, Han, Luke

**Main Antagonists:** Darth Vader, Boba Fett, Emperor

## Model Plot

```{r}
#| label: star wars
star_wars <- model_movie_example("Star Wars.*The Empire Strikes Back")
```

```{r}
#| label: plot star wars model
plot(star_wars$model)
par(mfrow = c(1, 1))
```

## Latent Positions Plot

```{r}
#| label: plot star wars graph
#| fig-width: 16
#| fig-height: 9
vertex.strength <- strength(star_wars$movie_graph)
plot(
    star_wars$movie_graph,
    edge.width = E(star_wars$movie_graph)$weight,
    vertex.size = vertex.strength / max(vertex.strength) * 20 + 10,
    layout = star_wars$model$U,
    vertex.label = stringr::str_replace_all(
        vertex_attr(star_wars$movie_graph)$name,
        "(\\w) (\\w)", 
        "\\1\n\\2"
    ),
    vertex.label.font = 2,
    vertex.label.cex = log(vertex.strength, 5) / max(log(vertex.strength, 3)) * 2,
    vertex.label.family = "sans",
    vertex.label.color = "black",
    mark.expand = 0,
    mark.border = NA,
)
```

# General Results

## Random Movie Sample

Took random sample of 20 movies and ran same procedure...
```{r}
#| label: read random movies sample
movies_sample <- read_tsv("488	The Last Samurai\n682	Reservoir Dogs\n130	Black Rain\n134	Blade Runner\n671	The Reader\n843	TRON\n495	Leviathan\n41	Alien: Resurrection\n816	Thirteen Days\n227	The Crying Game\n511	Lord of Illusions\n791	Sugar & Spice\n487	The Last of the Mohicans\n217	Crank\n116	Being Human\n471	The King's Speech\n189	Chasing Sleep\n244	Death to Smoochy\n320	French Kiss\n214	Coraline", col_names = c("id", "title"))

movies_sample$title
```

```{r}
#| label: loop over sample, write each
gof_stats_list = lapply(movies_sample$title, function(title) {
    out = model_movie_example(title)
    model = out$model

    gof = model$GOF |> data.frame() |> _[-1,] |> colMeans()

    return(gof)
})

observed_gof_stats_list = lapply(movies_sample$title, function(title) {
    out = model_movie_example(title)
    model = out$model

    gof = model$GOF |> data.frame()

    return(gof[1,])
})

model_mean_gof_stats = do.call(rbind, gof_stats_list) |> as.data.frame()

observed_gof_stats = do.call(rbind, observed_gof_stats_list) |> as.data.frame()

rownames(model_mean_gof_stats) = rownames(observed_gof_stats) = movies_sample$title

original_colnames = colnames(model_mean_gof_stats)
colnames(model_mean_gof_stats) = paste(original_colnames, "modeled", sep = ".")

colnames(observed_gof_stats) = paste(original_colnames, "observed", sep = ".")

comb_gof_stats = cbind(model_mean_gof_stats, observed_gof_stats)
```

## Goodness of Fits Residuals
```{r}
#| label: gof boxplots
gof_stats_residuals = (model_mean_gof_stats - observed_gof_stats) |> setNames(original_colnames) |> pivot_longer(cols = everything(), names_to = "GOF_Stat", values_to = "Residual")

ggplot(gof_stats_residuals, aes(x = GOF_Stat, y = Residual, fill = GOF_Stat)) +
    geom_boxplot() +
    theme_bw()
```

# Observed vs Residual
```{r}
#| label: base gof plot
p = ggplot(comb_gof_stats) +
    geom_abline(slope = 1, intercept = 0, color = "red") +
    theme_bw()
```

## sd(rowmean)
```{r}
#| label: sd.rowmean
p + geom_point(aes(x = sd.rowmean.observed, y = sd.rowmean.modeled))
```

## sd(colmean)
```{r}
#| label: sd.colmean
p + geom_point(aes(x = sd.colmean.observed, y = sd.colmean.modeled))
```

## dyad.dep
```{r}
#| label: dyad.dep
p + geom_point(aes(x = dyad.dep.observed, y = dyad.dep.modeled))
```

## cycle.dep
```{r}
#| label: cycle.dep
p + geom_point(aes(x = cycle.dep.observed, y = cycle.dep.modeled))
```

## trans.dep
```{r}
#| label: trans.dep
p + geom_point(aes(x = trans.dep.observed, y = trans.dep.modeled))
```