```{r, warning=FALSE, message=FALSE, results='hide'}
# Your code here will run, but warnings and text output won't show.
```

```{r}
#| label: sbm setup
#| include: false

#Fitting the Data using SBM with a Variational EM Method
#Importing Libraries potentially needed to fit the SBM with VEM 
library(tidyverse)
library(clue)
library(igraph)
library(netrankr)
library(blockmodels)
library(ggplot2)
library(reshape2)
library(dplyr)
library(rgexf)
library(sbm)
library(gridExtra)
library(patchwork)
library(ggrepel)

#movies_sample contains our specific sample of 20 movies

#Functions to generate a igraph objects of movies (referred from ame.qmd from Tony's Slides)
movies_metadata <- read_tsv("network_metadata.tab")

get_movie_id_by_title <- function(title) {
    movies_metadata |>
        filter(grepl(title, Title)) |>
        pull(GexfID)
}
```

# Stochastic Block Models and Variational EM Algorithm

We will be now using Stochastic Block Models to not only explore the community structure of our data but also get a more accurate model fit by accounting for the community structure in the model.

# Stochastic Block Models and Variational EM Algorithm

Since the Maximum Likelihood Estimate for our community assignment in not computable in polynomial time, we will be using the Variational EM Method where we consider a candidate variational distribution for Z, $q(Z) = \Pi_{i} \Pi_{q} \tau_{iq}^{z_{iq}}$ and iteratively optimize a objective function $F(q,\theta)$ over the E-step (variational parameters) and the M-step (model parameters):

$$
F(q,\theta) = E_{q(Z)}[log P(A,Z;\theta)] - E_{q}[log(q(Z))]
$$

where Z is our matrix of community assignment vectors (independent and indentically samples from a multinomial distribution) and $A_{ij}$ are observed data where $A_{ij}|(Z_iq=1,Z_{jl}=1) = Bernoulli(\pi_{ql})$

# Fitting SBM with Variational EM Method in R

We will use the `blockmodels` library, as we did in class, to fit our `moviegalaxies` data in R using the `BM_Bernoulli` function. Model Selection will be done by default using the Integrated Complete Likelihood Criteria (ICL) which will essentially select the optimal number of clusters/communities by balancing model fit with high clustering property.


```{r}
#| label: sbm_model_movie
#Functions to fit SBM Models with VEM (Template referring Tony's work in the ame.qmd slides)
dcsbm_model_movie <- function(title) {
    #Get the movie igraph object
    movie_id <- get_movie_id_by_title(title)
    movie_graph <- read.gexf(str_glue("data/{movie_id}.gexf")) |> gexf.to.igraph()
    
    #Getting the adjacency matrix for movie graph object 
    movieA <- as_adjacency_matrix(movie_graph, sparse=FALSE)
    
     # Symmetrize (just in case) and zero diagonal
    movieA <- (movieA + t(movieA)) / 2
    diag(movieA) <- 0

    # Bernoulli SBM with degree_correction
    model <- estimateSimpleSBM(
    netMat       = (movieA > 0) * 1,
    model        = "bernoulli",
    directed     = FALSE,
    estimOptions = list(
    verbosity     = 0,
    nbCores       = 1,
    degreeCorrect = TRUE))
    
    out <- list(movie_graph = movie_graph, model = model)
    return(out)
}

# Fit DCSBM for every movie
stopifnot(nrow(movies_sample) == 20)

results <- vector("list", nrow(movies_sample)) 
names(results) <- movies_sample$title

for (i in seq_along(movies_sample$title)) {
  title <- movies_sample$title[i]
  cat(sprintf("[%d/20] Fitting: %s\n", i, title))
  suppressMessages(
    results[[title]] <- dcsbm_model_movie(title)
  )
}
```


## SBM Model Info Collection

In case of a single network, Stochastic Block Models are powerful tools to analyze, model and further gain intuition and understanding about the community structure of the graph and analyzing interactions and gateways between communities. SBMs can also model the data for a Network and possess decent predictive power (even more by accounting for degree heterogeneity using DCSBM). 

In our case we are fitting multiple SBM models, which let's us instead to a comparative analysis among different movies using the `sbm` package. We will extract the metadata about each model fit using R:

```{r}
#Pass models to the function to get metadata
dcsbm_metadata <- function(res) {
  model <- res$model
  g <- res$movie_graph
  
  list(
    K         = model$nbBlocks, #Number of communities
    labels    = model$memberships, #assignment vectors
    block_mat = model$connectParam$mean,   #Block Probability matrix
    ICL       = model$ICL,     #ICL value of each model
    n_nodes   = vcount(g), #Num Nodes
    n_edges   = ecount(g), #Num edges
    graph     = g #graphs
  )
}

best <- lapply(results, dcsbm_metadata)
```


## Summarizing SBM Model Information

Summary of the DCSBM model information for all 20 movies:

```{r}
summary_df <- data.frame(
  movie   = names(best), 
  n_nodes= sapply(best, `[[`, "n_nodes"),
  n_edges = sapply(best, `[[`, "n_edges"),
  K_hat= sapply(best, `[[`, "K"),
  ICL = round(sapply(best, `[[`, "ICL"),2),
  modularity = round(mapply(function(b) {
    modularity(b$graph, b$labels)
  }, best),2)
)
```


```{r}
# --- Plot 1: Network Size - use pivot_longer but plot separately to avoid dodge issues
p1 <- ggplot(summary_df, aes(x = reorder(movie, n_nodes + n_edges))) +
  geom_col(aes(y = n_edges), fill = "coral", alpha = 0.8) +
  geom_col(aes(y = n_nodes), fill = "steelblue", alpha = 0.8) +
  coord_flip() +
  labs(title = "Network Size by Movie", x = "", y = "Count") +
  theme_minimal(base_size = 7)

# --- Plot 2: Optimal K
p2 <- ggplot(summary_df, aes(x = reorder(movie, K_hat),
                               y = K_hat,
                               fill = factor(K_hat))) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(breaks = 1:max(summary_df$K_hat)) +
  labs(title = "Optimal K by Movie", x = "", y = "K", fill = "K") +
  theme_minimal(base_size = 7)

# --- Plot 3: ICL
p3 <- ggplot(summary_df, aes(x = reorder(movie, ICL),
                               y = ICL,
                               fill = factor(K_hat))) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "ICL by Movie", x = "", y = "ICL", fill = "K") +
  theme_minimal(base_size = 7)

# --- Plot 4: Modularity
p4 <- ggplot(summary_df, aes(x = reorder(movie, modularity),
                               y = modularity,
                               fill = factor(K_hat))) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Modularity by Movie", x = "", y = "Modularity", fill = "K") +
  theme_minimal(base_size = 7)

(p1 + p2) / (p3 + p4)
```


##Homophily within characters' communities in the movies

```{r}
assort_score <- function(nm) {
  bmat         <- best[[nm]]$block_mat
  within_mean  <- mean(diag(bmat))
  between_mean <- mean(bmat[upper.tri(bmat)])
  if (between_mean == 0) return(NA)
  within_mean / between_mean
}

summary_df$assort_score <- sapply(names(best), assort_score)

ggplot(summary_df, aes(x = reorder(movie, assort_score), 
                        y = assort_score,
                        fill = factor(K_hat))) +
  geom_col() +
  geom_hline(yintercept = 1, linetype = "dashed", color = "black") +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title    = "Homophily Score by Movie",
    x        = "",
    y        = "Homophily Score",
    fill     = "Number of\nCommunities (K)"
  ) +
  theme_minimal()
```







## Organized v/s Core-Periphery Character World Building

- **X-axis (Homophily):** Within-block connection preference. Line at 1 = theoretical null (no preference).
- **Y-axis (Core-Periphery):** Variance in block connectivity. Line at sample median (var = 0 unrealistic for real networks).

```{r}
# ============================================================
# ANALYSIS 4: Interaction Grammar Scatter Plot
# ============================================================

compute_grammar_measures <- function(nm) {
  bmat <- best[[nm]]$block_mat
  K    <- nrow(bmat)
  
  if (K == 1) return(data.frame(movie = nm, homophily = NA, core_periphery = NA))
  
  # Homophily: diagonal mean / overall mean
  # Reference at 1 = theoretical null of no within-block preference
  diag_mean    <- mean(diag(bmat))
  overall_mean <- mean(bmat)
  homophily    <- ifelse(overall_mean == 0, NA, diag_mean / overall_mean)
  
  # Core-Periphery: variance of row means
  # Reference at sample median — variance = 0 is too idealistic
  # for real networks so we compare relative to sample
  row_means      <- rowMeans(bmat)
  core_periphery <- var(row_means)
  
  data.frame(
    movie          = nm,
    homophily      = homophily,
    core_periphery = core_periphery
  )
}

grammar_df <- do.call(rbind, lapply(names(best), compute_grammar_measures))
grammar_df <- merge(grammar_df, summary_df[, c("movie", "K_hat")], by = "movie")

# Theoretically motivated x reference (homophily = 1)
# Empirically motivated y reference (sample median)
cp_median <- median(grammar_df$core_periphery, na.rm = TRUE)

ggplot(grammar_df, aes(x = homophily,
                        y = core_periphery,
                        color = factor(K_hat),
                        label = movie)) +
  
  # Quadrant shading
  annotate("rect", xmin = -Inf, xmax = 1,
           ymin = cp_median, ymax = Inf,
           fill = "lightblue", alpha = 0.15) +
  annotate("rect", xmin = 1, xmax = Inf,
           ymin = cp_median, ymax = Inf,
           fill = "lightgreen", alpha = 0.15) +
  annotate("rect", xmin = -Inf, xmax = 1,
           ymin = -Inf, ymax = cp_median,
           fill = "lightyellow", alpha = 0.15) +
  annotate("rect", xmin = 1, xmax = Inf,
           ymin = -Inf, ymax = cp_median,
           fill = "lightsalmon", alpha = 0.15) +
  
  # Quadrant labels
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5, size = 2.8,
           color = "gray40", label = "Disassortative\nCore-Periphery") +
  annotate("text", x = Inf,  y = Inf, hjust = 1.1,  vjust = 1.5, size = 2.8,
           color = "gray40", label = "Assortative\nCore-Periphery") +
  annotate("text", x = -Inf, y = -Inf, hjust = -0.1, vjust = -0.5, size = 2.8,
           color = "gray40", label = "Disassortative\nDiffuse") +
  annotate("text", x = Inf,  y = -Inf, hjust = 1.1,  vjust = -0.5, size = 2.8,
           color = "gray40", label = "Assortative\nDiffuse") +
  
  # Reference lines
  geom_vline(xintercept = 1,
             linetype = "dashed", color = "gray60", linewidth = 0.4) +
  geom_hline(yintercept = cp_median,
             linetype = "dashed", color = "gray60", linewidth = 0.4) +
  
  geom_point(size = 3, alpha = 0.85) +
  ggrepel::geom_text_repel(size = 2.2, max.overlaps = 20) +
  
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Movie Interaction Grammar: Homophily vs Core-Periphery",
    x     = "Homophily  (within-block preference)",
    y     = "Core-Periphery  (variance in block connectivity)",
    color = "K"
  ) +
  theme_minimal(base_size = 10)
```

## Plots for Degree Heterogeneity

```{r}
# ============================================================
# ANALYSIS 5: Degree Heterogeneity within Movie Networks
#
# CV of degree measures how unequal character connectivity is.
# High CV => a few characters dominate most interactions
#            (star-driven narrative, clear protagonist)
# Low CV  => characters share interactions relatively equally
#            (ensemble cast, distributed narrative)
#
# We also compute CV within each SBM block to ask:
# does degree heterogeneity exist WITHIN communities or
# only ACROSS them? This directly justifies the DC-SBM —
# if CV within blocks is high, plain SBM would have split
# blocks purely by degree rather than narrative role.
# ============================================================

compute_degree_het <- function(nm) {
  g <- best[[nm]]$graph
  labels <- best[[nm]]$labels
  K <- best[[nm]]$K
  
  deg <- degree(g)
  
  # Global Normalized Deg Heterogenity across entire network
  global_cv <- sd(deg) / mean(deg)
  
  # Deg Heterogeniety within each block
  block_cvs <- sapply(1:K, function(k) {
    block_deg <- deg[labels == k]
    if (length(block_deg) < 2) return(NA)
    sd(block_deg) / mean(block_deg)
  })
  
  # Mean within-block CV (average heterogeneity inside communities)
  mean_within_cv <- mean(block_cvs, na.rm = TRUE)
  
  data.frame(
    movie          = nm,
    global_cv      = global_cv,
    mean_within_cv = mean_within_cv,
    K              = K
  )
}

het_df <- do.call(rbind, lapply(names(best), compute_degree_het))
het_df <- merge(het_df, summary_df[, c("movie", "K_hat")], by = "movie")

p_global <- ggplot(het_df, aes(x = reorder(movie, global_cv),
                                y = global_cv,
                                fill = factor(K_hat))) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Global Degree Heterogeneity by Movie",
    x     = "",
    y     = "CV of Degree",
    fill  = "K"
  ) +
  theme_minimal(base_size = 8) +
  theme(axis.text.y = element_text(size = 6))

p_within <- ggplot(het_df, aes(x = reorder(movie, mean_within_cv),
                                y = mean_within_cv,
                                fill = factor(K_hat))) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Within-Block Degree Heterogeneity by Movie",
    x     = "",
    y     = "Mean Within-Block CV of Degree",
    fill  = "K"
  ) +
  theme_minimal(base_size = 8) +
  theme(axis.text.y = element_text(size = 6))

# ── Plot 3: Global vs Within-block CV scatter ─────────────────────────────────
# Movies above the diagonal have MORE within-block heterogeneity
# relative to global — degree variation is not explained by
# block membership alone, strongly justifying DC-SBM
p_scatter <- ggplot(het_df, aes(x = global_cv,
                                 y = mean_within_cv,
                                 color = factor(K_hat),
                                 label = movie)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "gray60") +
  geom_point(size = 3, alpha = 0.85) +
  ggrepel::geom_text_repel(size = 2.2, max.overlaps = 20) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title    = "Global vs Within-Block Degree Heterogeneity",
    x        = "Global Degree Heterogeneity",
    y        = "Mean Within-Block Degree Heterogeneity",
    color    = "K"
  ) +
  theme(axis.text.y = element_text(size = 4))

# ── Print ─────────────────────────────────────────────────────────────────────
(p_global + p_within) / p_scatter
```
