# SBM Model

```{r}
#| label: sbm setup
#| include: false

#Fitting the Data using SBM with a Variational EM Method
#Importing Libraries potentially needed to fit the SBM with VEM 
library(tidyverse)
library(clue)
library(igraph)
library(netrankr)
library(blockmodels)
library(blockmodels)
library(ggplot2)
library(reshape2)
library(dplyr)
library(rgexf)

#movies_sample contains our specific sample of 20 movies

#Functions to generate a igraph objects of movies (referred from ame.qmd from Tony's Slides)
movies_metadata <- read_tsv("network_metadata.tab")

get_movie_id_by_title <- function(title) {
    movies_metadata |>
        filter(grepl(title, Title)) |>
        pull(GexfID)
}
```

# Stochastic Block Models and Variational EM Algorithm

We will be now using Stochastic Block Models to not only explore the community structure of our data but also get a more accurate model fit by accounting for the community structure in the model.

# Stochastic Block Models and Variational EM Algorithm

Since the Maximum Likelihood Estimate for our community assignment in not computable in polynomial time, we will be using the Variational EM Method where we consider a candidate variational distribution for Z, $q(Z) = \Pi_{i} \Pi_{q} \tau_{iq}^{z_{iq}}$ and iteratively optimize a objective function $F(q,\theta)$ over the E-step (variational parameters) and the M-step (model parameters):

$$
F(q,\theta) = E_{q(Z)}[log P(A,Z;\theta)] - E_{q}[log(q(Z))]
$$

where Z is our matrix of community assignment vectors (independent and indentically samples from a multinomial distribution) and $A_{ij}$ are observed data where $A_{ij}|(Z_iq=1,Z_{jl}=1) = Bernoulli(\pi_{ql})$

# Fitting SBM with Variational EM Method in R

We will use the `blockmodels` library, as we did in class, to fit our `moviegalaxies` data in R using the `BM_Bernoulli` function. Model Selection will be done by default using the Integrated Complete Likelihood Criteria (ICL) which will essentially select the optimal number of clusters/communities by balancing model fit with high clustering property.


```{r}
#| label: sbm_model_movie
#Functions to fit SBM Models with VEM (Template referring Tony's work in the ame.qmd slides)
dcsbm_model_movie <- function(title) {
    #Get the movie igraph object
    movie_id <- get_movie_id_by_title(title)
    movie_graph <- read.gexf(str_glue("data/{movie_id}.gexf")) |> gexf.to.igraph()
    
    #Getting the adjacency matrix for movie graph object 
    movieA <- as_adjacency_matrix(movie_graph, sparse=FALSE)
    
     # Symmetrize (just in case) and zero diagonal
    movieA <- (movieA + t(movieA)) / 2
    diag(movieA) <- 0

    # BM_bernoulli with degree_correction = TRUE  =>  DC-SBM
    # blockmodels uses VEM internally
    model <- BM_bernoulli(
      membership_type = "SBM_sym",
      adj             = (movieA > 0) * 1,
      verbosity       = 0,
      plotting        = ""
    )
    model$estimate()  # runs Variational EM for K=1..max_blocks
    out <- list(movie_graph = movie_graph, model = model)
    return(out)
}

# Fit DCSBM for every movie
stopifnot(nrow(movies_sample) == 20)

results <- vector("list", nrow(movies_sample)) 
names(results) <- movies_sample$title

for (i in seq_along(movies_sample$title)) {
  title <- movies_sample$title[i]
  cat(sprintf("[%d/20] Fitting: %s\n", i, title))
  suppressMessages(
    results[[title]] <- dcsbm_model_movie(title)
  )
}
```


# Analysis using the Stochastic Block Model Fits of our movies

In case of a single network, Stochastic Block Models are powerful tools to analyze, model and further gain intuition and understanding about the community structure of the graph and analyzing interactions and gateways between communities. SBMs can also model the data for a Network and possess decent predictive power (even more by accounting for degree heterogeneity using DCSBM). 

In our case we are fitting multiple SBM models, which let's us instead to a comparative analysis among different movies.


