```{r}
#| label: setup
#| include: false
base_path <- "data/"

movie_ids <- c(488, 682, 130, 134, 671, 843, 495, 41, 816, 227,
               511, 791, 487, 217, 116, 471, 189, 244, 320, 214)

movie_names <- c(
  "488" = "The Last Samurai",
  "682" = "Reservoir Dogs",
  "130" = "Black Rain",
  "134" = "Blade Runner",
  "671" = "The Reader",
  "843" = "TRON",
  "495" = "Leviathan",
  "41"  = "Alien: Resurrection",
  "816" = "Thirteen Days",
  "227" = "The Crying Game",
  "511" = "Lord of Illusions",
  "791" = "Sugar & Spice",
  "487" = "The Last of the Mohicans",
  "217" = "Crank",
  "116" = "Being Human",
  "471" = "The King's Speech",
  "189" = "Chasing Sleep",
  "244" = "Death to Smoochy",
  "320" = "French Kiss",
  "214" = "Coraline"
)

movie_igraphs <- list()
for (id in movie_ids) {
  file <- file.path(base_path, paste0(id, ".gexf"))
  if (file.exists(file)) {
    gexf_obj <- read.gexf(file)
    g <- graph_from_data_frame(
      d = gexf_obj$edges[, c("source", "target")],
      directed = FALSE,
      vertices = gexf_obj$nodes[, "id", drop = FALSE]
    )
    movie_igraphs[[as.character(id)]] <- g
  }
}

spectral_stats <- data.frame(
  movie_id = integer(), movie_name = character(),
  n_nodes = integer(), n_edges = integer(),
  n_components = integer(), lambda2 = numeric(),
  spectral_gap = numeric(), largest_eigenval = numeric(),
  spectral_spread = numeric()
)
eigenvalue_list <- list()

for (id in names(movie_igraphs)) {
  g <- movie_igraphs[[id]]
  g <- as.undirected(g, mode = "collapse")
  g <- simplify(g)
  g <- delete_vertices(g, which(degree(g) == 0))
  n <- vcount(g); m <- ecount(g)
  if (n < 3) next
  L <- laplacian_matrix(g, normalized = FALSE)
  eigs <- tryCatch(sort(eigen(as.matrix(L), symmetric = TRUE)$values), error = function(e) NULL)
  if (is.null(eigs) || any(!is.finite(eigs))) next
  eigenvalue_list[[id]] <- eigs
  spectral_stats <- rbind(spectral_stats, data.frame(
    movie_id         = as.integer(id),
    movie_name       = movie_names[id],
    n_nodes          = n,
    n_edges          = m,
    n_components     = sum(eigs < 1e-8),
    lambda2          = round(eigs[2], 4),
    spectral_gap     = round(eigs[2] - eigs[1], 4),
    largest_eigenval = round(eigs[n], 4),
    spectral_spread  = round(eigs[n] - eigs[2], 4)
  ))
}

n_points <- 100
spectrum_matrix <- t(sapply(eigenvalue_list, function(eigs) {
  approx(seq(0, 1, length.out = length(eigs)), eigs,
         xout = seq(0, 1, length.out = n_points))$y
}))
rownames(spectrum_matrix) <- movie_names[rownames(spectrum_matrix)]

dist_matrix <- as.matrix(dist(spectrum_matrix))
sigma <- median(dist_matrix)
similarity_matrix <- exp(-dist_matrix^2 / (2 * sigma^2))
diag(similarity_matrix) <- 0

D_sim <- diag(rowSums(similarity_matrix))
L_sim <- D_sim - similarity_matrix
D_inv_sqrt <- diag(1 / sqrt(diag(D_sim)))
L_norm <- D_inv_sqrt %*% L_sim %*% D_inv_sqrt

eigs_sim <- sort(eigen(L_norm, symmetric = TRUE)$values)
gaps <- diff(eigs_sim[1:10])
k <- which.max(gaps) + 1

eig_decomp <- eigen(L_norm, symmetric = TRUE)
embedding <- eig_decomp$vectors[, (20 - k + 1):20]
embedding <- embedding / sqrt(rowSums(embedding^2))

set.seed(42)
km <- kmeans(embedding, centers = k, nstart = 50)
clusters <- km$cluster
names(clusters) <- movie_names[names(eigenvalue_list)]
cluster_colors <- c("#E24B4A", "#1D9E75")

spectral_stats$cluster <- clusters[spectral_stats$movie_name]
```

## Spectral Analysis

Using the same random sample of 20 films, we then applied spectral analysis by treating each character network as a mathematical object and extracting structural properties from its eigenvalues.

::::: columns
::: {.column width="50%"}
**Our goal:**

To determine if the mathematical fingerprint of a film's character network reveals something about how the film is structured — without knowing anything about plot or genre?
:::

::: {.column width="50%"}
Our conclusion:

**The answer**

Yes. Spectral analysis found two distinct structural types : confined ensemble films vs protagonist-driven films.
:::
:::::

## Finding 1: Smaller casts aren't always less connected

```{r}
#| echo: false
#| fig-height: 5

ordered_idx   <- order(spectral_stats$lambda2)
ordered_names <- spectral_stats$movie_name[ordered_idx]
bar_colors    <- ifelse(spectral_stats$lambda2[ordered_idx] >= 1.0, "#E24B4A",
                 ifelse(spectral_stats$lambda2[ordered_idx] >= 0.7, "#EF9F27", "#378ADD"))

par(mar = c(8, 4, 2, 2), bg = NA)
barplot(
  spectral_stats$lambda2[ordered_idx],
  names.arg = ordered_names,
  las = 2, col = bar_colors, border = NA,
  ylab = "Fiedler Value (λ_2)",
  cex.names = 0.65, cex.axis = 0.85
)
abline(h = 1.0, lty = 2, col = "#999999", lwd = 1)
```

Leviathan (12 characters) is **7× more cohesive** than Blade Runner (25 characters). Network density matters far more than size.

## Finding 2: Larger casts are systematically less cohesive

```{r}
#| echo: false
#| fig-height: 4.5

par(mar = c(5, 4, 2, 2), bg = NA)
plot(spectral_stats$n_nodes, spectral_stats$lambda2,
     pch = 19, col = "#378ADD", cex = 1.4,
     xlab = "Number of characters (nodes)",
     ylab = "λ_2 (Fiedler Value)",
     cex.axis = 0.85, cex.lab = 0.9)
text(spectral_stats$n_nodes, spectral_stats$lambda2,
     labels = spectral_stats$movie_name,
     pos = 3, cex = 0.52, col = "#555555")
abline(lm(lambda2 ~ n_nodes, data = spectral_stats),
       col = "#E24B4A", lty = 2, lwd = 1.5)
```

As cast grows, characters can no longer all interact. The outliers above the line (Mohicans, Coraline, King's Speech) still have tight narrative focus despite larger casts.

The red trend line shows the negative relationship clearly. Spend time on the outliers — they're more interesting than the trend itself. Films above the line are "punching above their weight" in cohesion terms.

## Finding 3: Two structural types emerge automatically

```{r}
#| echo: false

pca <- prcomp(embedding)
pc  <- pca$x[, 1:2]
rownames(pc) <- movie_names[names(eigenvalue_list)]

par(mar = c(5, 4, 2, 2), bg = NA)
plot(pc[, 1], pc[, 2],
     col = cluster_colors[clusters],
     pch = 19, cex = 2,
     xlab = "PC1 (network density)", ylab = "PC2",
     cex.axis = 0.85, cex.lab = 0.9)
text(pc[, 1], pc[, 2],
     labels = rownames(pc),
     pos = 3, cex = 0.58,
     col = cluster_colors[clusters])
legend("bottomright",
       legend = c("Cluster 1 — dense ensembles",
                  "Cluster 2 — protagonist-driven"),
       col = cluster_colors, pch = 19, cex = 0.8,
       bty = "n")
```

## The two clusters

::::: columns
::: {.column width="50%"}
**Cluster 1 — Dense ensembles**

*Avg 24 nodes · 99 edges · λ₂ = 1.18*

-   Reservoir Dogs
-   TRON
-   Leviathan
-   Alien: Resurrection
-   The Last of the Mohicans
-   The King's Speech

Small casts where nearly every character interacts with nearly every other.
:::

::: {.column width="50%"}
**Cluster 2 — Protagonist-driven**

*Avg 36 nodes · 89 edges · λ₂ = 0.63*

-   The Last Samurai
-   Black Rain · Blade Runner
-   The Reader · Thirteen Days
-   The Crying Game · Crank
-   French Kiss · Coraline
-   
    -   5 more

Larger casts organized around a central character. Peripheral characters don't meet each other.
:::
:::::

## Potential reasons for density

::::: columns
::: {.column width="50%"}
**Why Cluster 1 is so dense**

The main setting of each movie: Reservoir Dogs (diner), Leviathan (submarine), Alien: Resurrection (spaceship), The King's Speech (therapy room).

These are all confined settings which force all characters to interact with each other

Confined settings force all characters to interact with each other and this is exactly what high edge density and high λ₂ capture.
:::

::: {.column width="50%"}
**Why Cluster 2 is sparser**

Protagonist-driven films introduce many characters who only interact with the hero and not with each other.

This creates a hub-and-spoke structure that spectral analysis detects immediately.
:::
:::::

## Summary

Here are our main 3 takeaways from the spectral analysis of the random sample of 20 movie character networks.

**1 — All networks are connected, but not equally** λ₂ ranges from 0.33 to 2.39 which is a 7× spread driven by setting and narrative style, not just cast size.

**2 — Size and density trade off** Larger casts consistently produce sparser networks. Films that break this trend have tight, focused narratives that keep all characters linked.

**3 — Two structural archetypes exist** Spectral clustering automatically finds a divide between dense ensemble films and protagonist-driven films. This matches real narrative structure and this can be discovered with no genre labels.
